\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-06-04 paper.qmd
Remove AIC calculations

The best model for the S-curve using UMAP (n_neighbors: $15$) was created with $420$ bins. It was determined to be the best model after removing edges with a length greater than $0.134$. The decision was made considering the AIC plot (@fig-mse-aic-scurve), which revealed the occurrence of the minimum at $420$. Additionally, visually, the model accurately captures the geometry of the S-curve by fitting the points everywhere (@fig-scurve_sc_best).

In contrast, a poor model for the S-curve with UMAP was created using only $190$ bins and removing edges with a length greater than $0.3$. This model is inadequate as it has a higher MSE compared to the best model. Visually, the model does not fit the points in some places and mismatches the pattern (@fig-scurve_sc_bad).

<!-- Find the best model-->
```{r}
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:25 #sqrt(NROW(training_data_scurve)/2)

  scurve_model <- fit_highd_model(
    training_data = training_data_scurve,
    emb_df = umap_scurve_scaled,
    bin1 = xbins,
    r2 = r2,
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = TRUE,
    col_start_highd = "x"
  )

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_scurve,
    df_bin = df_bin_scurve,
    training_data = training_data_scurve,
    newdata = NULL,
    type_NLDR = "UMAP",
    col_start = "x") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_scurve),
           aic = 2*(NCOL(training_data_scurve)-1)*b + 2*NROW(training_data_scurve)*(NCOL(training_data_scurve)-1)*log(MSE))

best_model_b <- error_scurve |>
                    filter(aic == min(aic)) |>
                    pull(b)

mse_scurve_b <- ggplot(error_scurve, 
                      aes(x = b, 
                          y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = best_model_b, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  ylab("log(MSE)") +
  xlab("b") +
  theme(axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90))

aic_scurve_b <- ggplot(error_scurve, 
                      aes(x = b, 
                          y = aic)) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = best_model_b, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  ylab("AIC") +
  xlab("b") +
  theme(axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90))

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent $b$ choices. What is the effective $b$ to create the model? The MSE plot have a steep slope at the beginning, indicating that a smaller $b$ causes a larger amount of MSE. Then, the slope gradually declines or level off, indicating that a higher $b$ generates a smaller MSE. Using the elbow method, it was observed that when the $b$ is set to $420$, the lowest MSE occurred."
#| label: fig-mse-aic-scurve
#| fig-pos: H
#| out-height: 50%
mse_scurve_b + aic_scurve_b
```

>>>>

The appropriate model for the S-curve using UMAP (n_neighbors: $15$) was created with $420$ bins. It was determined to be the best model after removing edges with a length greater than $0.134$. The decision was made considering that the model accurately captures the geometry of the S-curve by fitting the points everywhere (@fig-scurve_sc_best).

In contrast, a poor model for the S-curve with UMAP was created using only $190$ bins and removing edges with a length greater than $0.3$. This model is inadequate as it has a higher MSE compared to the appropriate model. Visually, the model does not fit the points in some places and mismatches the pattern (@fig-scurve_sc_bad).

### 2024-06-04 paper.tex
Remove AIC calculations

The best model for the S-curve using UMAP (n\_neighbors: \(15\)) was
created with \(420\) bins. It was determined to be the best model after
removing edges with a length greater than \(0.134\). The decision was
made considering the AIC plot (Figure~\ref{fig-mse-aic-scurve}), which
revealed the occurrence of the minimum at \(420\). Additionally,
visually, the model accurately captures the geometry of the S-curve by
fitting the points everywhere (Figure~\ref{fig-scurve_sc_best}).

In contrast, a poor model for the S-curve with UMAP was created using
only \(190\) bins and removing edges with a length greater than \(0.3\).
This model is inadequate as it has a higher MSE compared to the best
model. Visually, the model does not fit the points in some places and
mismatches the pattern (Figure~\ref{fig-scurve_sc_bad}).

\includegraphics[width=1\textwidth,height=0.5\textheight]{paper_files/figure-pdf/fig-mse-aic-scurve-1.pdf}

\caption{\label{fig-mse-aic-scurve}MSE from UMAP applied to S-curve
dataset with diffferent \(b\) choices. What is the effective \(b\) to
create the model? The MSE plot have a steep slope at the beginning,
indicating that a smaller \(b\) causes a larger amount of MSE. Then, the
slope gradually declines or level off, indicating that a higher \(b\)
generates a smaller MSE. Using the elbow method, it was observed that
when the \(b\) is set to \(420\), the lowest MSE occurred.}

>>>>

The appropriate model for the S-curve using UMAP (n\_neighbors: \(15\))
was created with \(420\) bins. It was determined to be the best model
after removing edges with a length greater than \(0.134\). The decision
was made considering that the model accurately captures the geometry of
the S-curve by fitting the points everywhere
(Figure~\ref{fig-scurve_sc_best}).

In contrast, a poor model for the S-curve with UMAP was created using
only \(190\) bins and removing edges with a length greater than \(0.3\).
This model is inadequate as it has a higher MSE compared to the
appropriate model. Visually, the model does not fit the points in some
places and mismatches the pattern (Figure~\ref{fig-scurve_sc_bad}).
